Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2018-01-22 2017-Xudong-SMT/FM-2018-extended.tex
section 4

\begin{abstract}
Open pNets are used to model the behavior of open systems, both
synchronous or asynchronous, expressed in various calculi or
languages. They are endowed with a symbolic operational semantics in
terms of so-called ``Open Automata''. This allows to check properties of
such systems in a compositional manner. 
We implement an algorithm computing this semantics,
building predicates expressing the synchronization conditions
between the events of the pNet sub-systems. Checking such
predicates requires symbolic reasoning over first order logics, but
also over application-specific data. We use the Z3 SMT engine to check
satisfiability of the predicates, and prune the open automaton of its
unsatisfiable transitions.
As an industrial oriented use-case, we use so-called "architectures"
of BIP systems, that have been used to specify the control software of
ESA microsatellites. We use pNets to encode BIP architectures,
extended with explicite data in synchronisations, and we compute their
open automaton semantics. This automaton may then be used to prove
behavioral properties, like e.g., \TODO{ par exemple ..}

\end{abstract}

@@

The semantics of open pNets will be defined as an open automaton, that
is an automaton where each transition composes transitions 
of several \noteSB{These should all be pLTSs, right? Or did I miss
  something?}\noteEM{Hum, nor really, something intermediate; the actions there can be
  parameterised actions, but the label has no guard and no assignments}\hl{LTSs} with the
actions of some holes; the transition occurs if some predicates hold,
and can involve a set of state modifications. Each state of an open
automaton has a set of \emph{state variables} that can be assigned by
incoming transitions.

\begin{definition}[Open transitions]
	\label{def:OpenTransitions}
	An \emph{open transition} over a set  $\langle S_i,s_{0 i}, \rightarrow_i\rangle^{i\in
	I}$ of \hl{LTSs}, a
	set $J$ of holes with sorts $Sort_j^{j\in J}$, and a set of states $\mathcal{S}$ is a structure of the form:	
	\begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
		\{\xrightarrow{b_j}_j\}^{j\in J}, \Pred\,, \Post}
	{s \xrightarrow {v}s'}\,,
	\end{mathpar}
	where $s, s'\in\mathcal{S}$ and, for all
        $i\in I$, $s_i{\xrightarrow{a_i}}_i s_i^{\prime}$ is a transition of the
	\hl{LTS} $\langle S_i,s_{0 i}, \rightarrow_i\rangle$, and $\xrightarrow{b_j}_j$
        is a transition of the hole $j$, for any action $b_j$ in the
        sort $Sort_j$. \Pred\ is a predicate
	over the variables of the terms, labels, and states $s_i$,
        $b_j$, $s$, $v$. \Post\ is a set of equations that  
	hold \emph{after the open transition}, represented as a
        substitution $\{x_k\gets e_k\}^{k\in K}$  
	where $x_k$ are variables of $s'$ and $s'_i$, whereas $e_k$ are
        expressions over the other variables of the open transition.
\end{definition}

\begin{definition}[Open automaton]
	\label{def:open-automaton}
	An \emph{open automaton} is a structure\\ $A =
	\langle \mathit{LTS}_i^{i\in I},J,\mathcal{S},s_0,\mathcal{T}\rangle$ where:
	\begin{itemize}
		\item[$\bullet$]  $I$ and $J$ are  sets of indices,
		\item[$\bullet$]  \noteSB{And one more above}\hl{$LTS_i^{i\in I}$} is a family of \hl{LTSs},
		\item[$\bullet$]   $\mathcal{S}$ is a set of states and $s_0 \in \mathcal{S}$ the initial state,
		\item[$\bullet$] $\mathcal{T}$ is a set of open transitions and, for each
		$t\in \mathcal{T}$, there exist $I'$, $J'$ with $I'\subseteq I$, $J'
		\subseteq J$, such that $t$ is an open transition over $LTS_i^{i\in I'}$, $J'$,
		and  $\mathcal{S}$.

@@

\noteSB{line 8: $E$ is never updated, so $E = \emptyset$ always;\\
line 11: Would it not be better to prune unsatisfiable transitions on-the-fly, \ie inside the loop?}
Algorithm~\ref{alg1} starts with an open pNet, and builds its set of open
transitions. Its main loop is a classical residual algorithm: starting
from the initial global state, it picks a state in an unexplored set, and
computes all possible OTs, adding their target states in the
unexplored set, until this set is empty. As the potential set of
global states is finite (see \cite{henrio:Forte2016}), this terminates.

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require The pNet node P.
\State Initialize sets $U=\mathit{InitState}(P)$, $E=\emptyset$ for unexplored/explored global states, $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$;
	\State $\mathit{OT} = \mathit{MakeTransitions}(P, S)$;
        \State Store $\mathit{OT}$ in $L$; 
	\ForAll{$\mathit{OT} \in L$}
        \State Let $T$ be the target global state of $\mathit{OT}$
	\IfThen{($T \not\in U$ \&\& \hl{$T \not\in E$})}{Add $T$ into $U$;}
	\EndFor
\EndWhile
\State \hl{Prune the open transitions in $L$ using the SMT solver;}
\State \Return $\mathit{OA}=\mathit{Reachable}(\mathit{InitState}(P),L)$;

@@

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
\hl{For pLTSs at the leaves, we simply take the pLTS transitions of the
corresponding local state and build the predicate to generate the
open transition.}
On a pNet node we use two methods, combining and
matching, to generate the predicate of the open transitions in a
hierarchical manner.
%
Then, at the root of the pNet, the predicate of each OT is translated
into SMTlib assertions, and checked for satisfiability. Unsatisfiable
OTs are pruned, and the final open automaton is computed as the set of
global states reachable by satisfiable OTs.

@@

\noteInSB{Line 2: Tr1 and Tr2 are the semantic rules, right?}

@@

\noteInSB{Line 8: When is $\mathit{Filter}(ot)$ called?}

\begin{algorithm}
\caption{MakeTransitions()}
  \label{alg2}
\begin{algorithmic}[1]
\Require a pNet node $Pn$; a global state $S$.
\State Initialize empty list $l$ and set $L$ for sub-transitions/transitions.
	\For{each $\mathit{Subnet}$ in $Pn$}
	~~~ $\setminus \setminus$ Recursively applying \hl{Tr1 or Tr2} on the sub-nodes.
                \State Store $\mathit{MakeTransitions}(Subnet, S)$ in $l$;
	\EndFor
	\For{each $sv \in SV$}
		\State $\overline{comb} = \mathit{Combining}(l)$;
                 \State \hl{$ot = \mathit{Matching}(sv, \overline{\mathit{comb}}, \mathit{hole})$;} % SB: removed the fourth parameter v here
                 \State Store $ot$ in $L$; ~~~ \hl{$\setminus \setminus$ if $\mathit{Filter}(ot)$ succeeds}
	\EndFor
\State \Return $L$;
\end{algorithmic}  
\end{algorithm}

@@

\def\inactive{\{-\}}
\paragraph{Combining.}
The combining method enumerates all the possible behaviours of the
subnets, as all the possible combinations of their open transitions.
Assume that there is a collection of $n$ subnets.\noteSB{These variables are not used and conflict with the returned set $L$ in Alg.~\ref{alg2}, on one hand, and control states $s_i$ of the subnets, on the other.}\st{
$L = {s_1, s_2, ...,  s_n}$;} We denote $\overline{ot}_i$ the set of open
transitions of the $i$-th
subnet; %% $s_i$ and 
``$-$'' means that the subnet is not involved.
The combination $\overline{comb}$, a set of n-tuples, is the
cartesian product:  
$\overline{comb} = (\inactive\cup \overline{ot}_1) \times (\inactive\cup \overline{ot}_2)\times \dots \times (\inactive\cup \overline{ot}_n)$\hl{, where, for each $i \in [1, n]$, $\overline{ot}_i$ is the set of open transitions obtained from the $i$-th subnet in line~3 of the algorithm}.

@@

\st{Each of these possible filters is more precise than the previous
ones. But e}Even when unification succeeds, it is still possible that
the resulting predicate would be unsatisfiable, because of some
incompatibility involving the guards. In our algorithm, we choose to
apply only the simplest filter inside the \emph{Matching} method. All
the rest will be checked later, using the satisfiability check in the
SMT engine.

>>>>

\begin{abstract}
Open pNets are used to model the behavior of open systems, both
synchronous or asynchronous, expressed in various calculi or
languages. They are endowed with a symbolic operational semantics in
terms of so-called ``Open Automata''. This allows to check properties of
such systems in a compositional manner. 
We implement an algorithm computing this semantics,
building predicates expressing the synchronization conditions
between the events of the pNet sub-systems. Checking such
predicates requires symbolic reasoning over first order logics, but
also over application-specific data. We use the Z3 SMT engine to check
satisfiability of the predicates, and prune the open automaton of its
unsatisfiable transitions.
As an industrial oriented use-case, we use so-called "architectures"
of BIP systems, that have been used to specify the control software of
ESA microsatellites. We use pNets to encode BIP architectures,
extended with explicite data in synchronisations, and we compute their
open automaton semantics. This automaton may then be used to prove
behavioral properties; we give 2 examples, a safety and a liveness
properties of a BIP architecture.
\end{abstract}

@@

The semantics of open pNets will be defined as an open automaton, that
is an automaton where each transition composes transitions 
of several
% \noteSB{These should all be pLTSs, right? Or did I miss
%  something?}\noteEM{Hum, nor really, something intermediate; the actions there can be
%  parameterised actions, but the label has no guard and no
%  assignments}
LTSs with the actions of some holes; the transition occurs if some predicates hold,
and can involve a set of state modifications. Each state of an open
automaton has a set of \emph{state variables} that can be assigned by
incoming transitions. Strictly speaking, the LTSs at the leaves of the
open automaton are a restricted form of pLTSs, where labels are
parametrised actions, but include no guard nor assignments.

\begin{definition}[Open transitions]
	\label{def:OpenTransitions}
	An \emph{open transition} over a set  $\langle S_i,s_{0 i}, \rightarrow_i\rangle^{i\in
	I}$ of LTSs, a
	set $J$ of holes with sorts $Sort_j^{j\in J}$, and a set of states $\mathcal{S}$ is a structure of the form:	
	\begin{mathpar}
	\inferrule*[myfraction=\reddottedrule]
	{\{s_i~{\xrightarrow{a_i}}_i ~s_i^{\prime}\}^{i\in I},
		\{\xrightarrow{b_j}_j\}^{j\in J}, \Pred\,, \Post}
	{s \xrightarrow {v}s'}\,,
	\end{mathpar}
	where $s, s'\in\mathcal{S}$ and, for all
        $i\in I$, $s_i{\xrightarrow{a_i}}_i s_i^{\prime}$ is a transition of the
	LTS $\langle S_i,s_{0 i}, \rightarrow_i\rangle$, and $\xrightarrow{b_j}_j$
        is a transition of the hole $j$, for any action $b_j$ in the
        sort $Sort_j$. \Pred\ is a predicate
	over the variables of the terms, labels, and states $s_i$,
        $b_j$, $s$, $v$. \Post\ is a set of equations that  
	hold \emph{after the open transition}, represented as a
        substitution $\{x_k\gets e_k\}^{k\in K}$  
	where $x_k$ are variables of $s'$ and $s'_i$, whereas $e_k$ are
        expressions over the other variables of the open transition.
\end{definition}

\begin{definition}[Open automaton]
	\label{def:open-automaton}
	An \emph{open automaton} is a structure\\ $A =
	\langle \mathit{LTS}_i^{i\in I},J,\mathcal{S},s_0,\mathcal{T}\rangle$ where:
	\begin{itemize}
		\item[$\bullet$]  $I$ and $J$ are  sets of indices,
		\item[$\bullet$]  $LTS_i^{i\in I}$ is a family of LTSs,
		\item[$\bullet$]  $\mathcal{S}$ is a set of states and $s_0 \in \mathcal{S}$ the initial state,
		\item[$\bullet$] $\mathcal{T}$ is a set of open transitions and, for each
		$t\in \mathcal{T}$, there exist $I'$, $J'$ with $I'\subseteq I$, $J'
		\subseteq J$, such that $t$ is an open transition over $LTS_i^{i\in I'}$, $J'$,
		and  $\mathcal{S}$.

@@

\noteSB{
line 11: Would it not be better to prune unsatisfiable transitions
on-the-fly, \ie inside the loop?}
\noteEM{Not so easy, it would require a proof... on one side this
  would implement reachability on the fly; but there is the question
  of collecting the assignments of the state variables, that are used
  in the sat check !}
Algorithm~\ref{alg1} starts with an open pNet, and builds its set of open
transitions. Its main loop is a classical residual algorithm: starting
from the initial global state, it picks a state in an unexplored set, and
computes all possible OTs, adding their target states in the
unexplored set, until this set is empty. As the potential set of
global states is finite (see \cite{henrio:Forte2016}), this terminates.

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require The pNet pn (cannot be reduce to a hole)
\State Initialize sets $U=\{\mathit{InitState}(pn)\}$, $E=\emptyset$
for unexplored/explored global states, $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; add $S$ in $E$;
	\State $\mathit{OT} = \mathit{MakeTransitions}(P, S)$;
        \State Store $\mathit{OT}$ in $L$; 
	\ForAll{$\mathit{OT} \in L$}
        \State Let $T$ be the target global state of $\mathit{OT}$
	\IfThen{($T \not\in U$ \&\& $T \not\in E$)}{Add $T$ into $U$;}
	\EndFor
\EndWhile
\State \hl{Prune the open transitions in $L$ using the SMT solver;}
\State \Return $\mathit{OA}=\mathit{Reachable}(\mathit{InitState}(pn),L)$;

@@

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
When applied to a pLTSs at the leaves, we simply take the pLTS transitions of the
corresponding local state and use the semantic rule (see TR1 in
Appendix~\ref{appendix:semRules}) to build the OT. 
When applied to a pNet node we use two methods, combining and
matching, to generate the open transitions in a hierarchical manner,
as detailed in algorithm~\ref{}. This method also manages directly the
holes of the node, so \emph{MakeTransitions} will never be called on a Hole.

%
Then, at the root of the pNet, the predicate of each OT is translated
into SMTlib assertions, and checked for satisfiability. Unsatisfiable
OTs are pruned, and the final open automaton is computed as the set of
global states reachable by satisfiable OTs.

@@

%\noteInSB{Line 2: Tr1 and Tr2 are the semantic rules, right?}

@@

%\noteInSB{Line 8: When is $\mathit{Filter}(ot)$ called?}

\begin{algorithm}
\caption{MakeTransitions() for pNet nodes}
  \label{alg2}
\begin{algorithmic}[1]
\Require a pNet node $Pn$; a global state $S$.
\State Initialize empty list $l$ and set $L$ for sub-transitions/transitions.
	\For{each $\mathit{Subnet}$ in $Pn$}
	~~~ $\setminus \setminus$ Recursively applying the semantic rules on the sub-nodes.
                \State Store $\mathit{MakeTransitions}(Subnet, S)$ in $l$;
	\EndFor
	\For{each $sv \in SV$}
		\State $\overline{comb} = \mathit{Combining}(l)$;
                 \State \hl{$ot = \mathit{Matching}(sv, \overline{\mathit{comb}}, \mathit{hole})$;} % SB: removed the fourth parameter v here
                 \State Store $ot$ in $L$; ~~~ \hl{$\setminus \setminus$ when $\mathit{Matching}(ot)$ succeeds}
	\EndFor
\State \Return $L$;
\end{algorithmic}  
\end{algorithm}

@@

\def\inactive{\{-\}}
\paragraph{Combining.}
The combining method enumerates all the possible behaviours of the
subnets, as all the possible combinations of their open transitions.
Assume that there is a collection of $n$ subnets.
%% \noteSB{These variables are not used and conflict with the returned
%%   set $L$ in Alg.~\ref{alg2}, on one hand, and control states $s_i$ of
%%   the subnets, on the other.}
%% \st{$L = {s_1, s_2, ...,  s_n}$;}
We denote $\overline{ot}_i$ the set of open
transitions of the $i$-th
subnet; %% $s_i$ and 
``$-$'' means that the subnet is not involved.
The combination $\overline{comb}$, a set of n-tuples, is the
cartesian product:  
$\overline{comb} = (\inactive\cup \overline{ot}_1) \times (\inactive\cup \overline{ot}_2)\times \dots \times (\inactive\cup \overline{ot}_n)$\hl{, where, for each $i \in [1, n]$, $\overline{ot}_i$ is the set of open transitions obtained from the $i$-th subnet in line~3 of the algorithm}.

@@

%\st{Each of these possible filters is more precise than the previous
%  ones. But e}
Even when unification succeeds, it is still possible that
the resulting predicate would be unsatisfiable, because of some
incompatibility involving the guards. In our algorithm, we choose to
apply only the simplest filter inside the \emph{Matching} method. All
the rest will be checked later, using the satisfiability check in the
SMT engine.
